---
title: "fhircrackr: Handling HL7 FHIR resources in R"

date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fhircrackr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
`fhircrackr` is a package designed to help analyzing HL7 FHIR resources.

FHIR stands for *Fast Healthcare Interoperability Resources* and  is a standard describing data formats and elements (known as "resources") as well as an application programming interface (API) for exchanging electronic health records. The standard was created by the Health Level Seven International (HL7) health-care standards organization. For more information on the FHIR standard, visit https://www.hl7.org/fhir/.

While FHIR is a very useful standard to describe and exchange medical data in an interoperable way, it is not very useful for statistical analyses of said data. This is due to the fact that FHIR data is stored in many nested and interlinked resources instead of matrix-like structures.

Thus, to be able to do statistical analyses a tool is needed that allows converting these nested resources into data frames. This process of flattening FHIR resources is not trivial, as the unpredictable degree of nesting and connectedness of the resources makes generic solutions to this problem not feasible.

We therefore implemented a package that makes it possible to download FHIR resources from a server into R and to flatten these resources into (multiple) data frames.

The package is still under development. The CRAN version of the package contains all functions that are already stable, for more recent (but potentially unstable) developments, the development version of the package can be downloaded from GitHub using `devtools::install_github("POLAR-fhiR/fhircrackr")`.

## Prerequisites
The complexity of the problem requires a couple of prerequisites both regarding knowledge and access to data. We will shortly list the  preconditions to using the `fhircrackr` package here:  

1. First of all, you need the endpoint of the fhir server you want to access. If you don't have you own fhir server, you can use one of the publicly available servers, such as [https://hapi.fhir.org/baseR4](https://hapi.fhir.org/baseR4) or [http://fhir.hl7.de:8080/baseDstu3](http://fhir.hl7.de:8080/baseDstu3). The endpoint of a fhir server is often referred to as [base].

2. To download ressources from the server, you should be familiar with [FHIR search requests](https://www.hl7.org/fhir/search.html). Fhir search allows you to download sets of resources that match very specific requirements. As the focus of this package is dealing with FHIR resources in R, rather than the intricacies of FHIR search, we will mostly use simple examples of FHIR search requests. Most of them will have the form `[base]/[type]?parameter(s)`, where `[type]` refers to the type of resource you are looking for and `parameter(s)` characterise specific properties those resources should have.
`http://hapi.fhir.org/baseR4/Patient?gender=female` for example downloads all Patient resources from the fhir server at `http://hapi.fhir.org/baseR4/` that represent female patients.

3. In the first step, `fhircrackr` downloads the resources in xml format into R. To specify which elements from the FHIR resources you want in your data frame, you should have at least some familiarity with XPath expressions. A good tutorial on XPath expressions can be found [here](https://www.w3schools.com/xml/xpath_intro.asp).

In the following we'll go through a typical workflow with `fhircrackr` step by step.

## Download and flatten fhir resources from a server

### Example 1: Download Patient resources
We will start with a very simple example and use `fhir_search()` to download Patient resources from a publicly available HAPI server after we've loaded the package with `library(fhircrackr)`:

```{r, results=F, message=F}
library(fhircrackr)
patient_bundles <- fhir_search(request="http://hapi.fhir.org/baseR4/Patient?", max.bundles=2)

```

The minimum information `fhir_search()` requires is a string containing the full FHIR search request in the argument `request`.
In general, a fhir search request returns a *bundle* of the resources you requested. If there are a lot of resources matching your request, the search result isn't returned in one big bundle but distributed over several of them. If the argument `max.bundles` is set to its default `Inf`, `fhir_search()` will return all available bundles, meaning all resources matching your request. If you set it to `2` as in the above example, the download will stop after the first two bundles. Note that in this case, the result *may not contain all* the resources from the server matching your request.

If you want to connect to a fhir server that uses basic authentification, you can supply the arguments `username` and `password`.

Because endpoints can sometimes be hard to reach, `fhir_search()` will start five attempts to connect to the endpoint before it gives up. With the arguments `max.attempts` and `delay.between.attempts` you can control this number as well the time interval between attempts.

As you can see in the next block of code, `fhir_search()` returns a list of xml objects where each list element represents one bundle of resources, so a list of two xml objects in our case:

```{r}
str(patient_bundles)
```


We know that inside these xml objects there is the patient data somewhere. To get it out, we will use `fhir_crack()`. The most important arguments `fhir_crack()` takes is `bundles`, the list of bundles that is returned by `fhir_search()` and `design`, an object that tells the function wich data to extract from the bundle. It returns a list of data.frames.

In general, `design` is a named list containing one element per data frame that will be created. The element names of `design` are going to be the names of the resulting data frames. It usually makes sense to create one data frame per type of resource. Because we have just downloaded resources of the type Patient, the `design` here would be a list of length 1. 

The elements of `design` are lists themselves, they can be of length 1 or length 2, depending on the level of precision in extracting the attributes. There are three levels of precision in extracting the data for our data frame with `fhir_crack()`:

#### 1. Extract all available attributes 
If we want to extract all available attributes, the list describing the data frames within design is a list of length 1, containing only an Xpath expression to the resource type we want to extract: 

```{r}
#define which elements of the resources are of interest
design1 <- list(
  Patients = list(
    ".//Patient"
  )
)

#Convert resources
list_of_tables <- fhir_crack(patient_bundles, design1)

#have look at part of the results
list_of_tables$Patients[1:5,1:6]
```

#### 2. Extract all attributes at a certain level
We can extract all attributes that are found on a certain level of the resource if we specify this level in an XPath expression and provide it as the second element of the list describing the data.frame:

```{r}
#define which elements of the resources are of interest
design2 <- list(
  Patients = list(
    ".//Patient",
    "./*/@value"
  )
)

#Convert resources
list_of_tables <- fhir_crack(patient_bundles, design2)

#have look at the results
head(list_of_tables$Patients)
```

`"./*/@value"` for example tells `fhir_crack()` to extract all attributes that are located one level below the root level.

#### 3. Extract specific attributes
If we know exactly which attributes to extract, we can specify them in a named list that we provide as the second element of the list describing the data.frame:

```{r}
#define which elements of the resources are of interest

#define which elements of the resources are of interest
design3 <- list(
	
	Patients = list(
		".//Patient",
		list(
			PID           = "id/@value",
			NAME.USE      = "name/use/@value",
			NAME.GIVEN    = "name/given/@value",
			NAME.FAMILY   = "name/family/@value",
			GENDER        = "gender/@value",
			BIRTHDAY      = "birthDate/@value"
		)
	)
) 
#Convert resources
list_of_tables <- fhir_crack(patient_bundles, design3)

#have look at the results
head(list_of_tables$Patients)
```

Of course these are examples for just one resource type. If you are interested in several types of resources, `design` will have more elements and the result will be a list of several data frames.

The abstract form `design` should therefore have is:

```{r, eval=F}
list(
	
#Option 1: extract all attributes	
  <Name of first data frame> = list(
    <XPath to resource type>
  ),

#Option 2: extract attributes from certain level  
  <Name of second data frame> = list(
    <XPath to resource type>,
    <XPath indicating attribute level> 
  ),

#Option 3: extrct specific attributes  
  <Name of third data frame> = list(
    <XPath to resource type>,
    list(
      <column name 1> = <XPath to attribute>
      <column name 2> = <XPath to attribute>
      ...
    ) 
  ),
  ...
)
```

### Example 2: Download MedicationStatement and corresponding Patient resources
In reality your FHIR search requests are probably going to be slightly more complex than just asking for Patient resources. Consider the following example where we want to download MedicationStatements refering to a certain medication we specify with its snomed code and also the Patient resources the MedicationStatements are linked to.

When the FHIR search request gets longer, it can be helpful to build up the request piece by piece like this:

```{r}
search_request  <- paste0(
  "https://hapi.fhir.org/baseR4/", #server endpoint
  "MedicationStatement?", #look for MedicationsStatements
  "code=http://snomed.info/ct|429374003", #only choose resources with this loinc code
  "&_include=MedicationStatement:subject") #include the corresponding Patient resources
```

Then we can download the resources:

```{r, results=F, message=F}
medication_bundles <- fhir_search(search_request, max.bundles = 3)
```

And convert them into to data frames, one for the MedicationStatements and one for the Patients:

```{r}
design <- list(

	MedicationStatement = list(

		".//MedicationStatement",

		list(
			AID                = "id/@value",
			STATUS.TEXT        ="text/status/@value",
			STATUS             = "status/@value",
			MEDICATION.SYSTEM  = "medicationCodeableConcept/coding/system/@value",
			MEDICATION.CODE    = "medicationCodeableConcept/coding/code/@value",
			MEDICATION.DISPLAY = "medicationCodeableConcept/coding/display/@value",
			DOSAGE             = "dosage/text/@value",
			PATIENT            = "subject/reference/@value",
			LAST.UPDATE        = "meta/lastUpdated/@value"
		)
	),

	Patients = list(

		".//Patient",
		"./*/@value"
	)
)


list_of_tables <- fhir_crack(medication_bundles, design)

head(list_of_tables$MedicationStatement)

head(list_of_tables$Patients)
```

## Saving and loading downloaded bundles
Since `fhir_crack()` discards of all the data not specified in `design` it makes sense to store the original search result for reproducibility and in case you realise later on that you need elements from the resources that you haven't extracted at first.

`fhir_save()` takes a list of bundles in form of xml objects (as returned by `fhir_search()`) and writes them into the directory specified in the argument `directory`. Each bundle is saved as a seperate xml-file. If the folder defined in `directory` doesn't exist, it is created in the current working directory.

```{r}
#save bundles as xml files
fhir_save(patient_bundles, directory="MyDirectory")
```

To read bundles saved with `fhir_save()` back into R, you can use `fhir_load()`:

```{r}
bundles <- fhir_load("MyDirectory")
```  

`fhir_load()` takes the name of the directory (or path to it) as its only argument. All xml-files in this directory will be read into R and returned as a list of bundles in xml format just as returned by `fhir_search()`.


## Other functions

### Download capability statement
The [capability statement](https://www.hl7.org/fhir/capabilitystatement.html) documents  a set of capabilities (behaviors) of a FHIR Server for a particular version of FHIR. You can download this statement using the function `fhir_cs()`:

```{r, results=F}
cap <- fhir_cs("http://hapi.fhir.org/baseR4/")
```

`fhir_sc()` takes a FHIR server endpoint and returns a list of data frames containing all information from the capability statement of this server.

You can then access the parts that interest you, for example:

```{r}
cap$META$software.version
```
