---
title: "fhircrackr"
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
`fhircrackr` is a package designed to help analyzing HL7 FHIR^[FHIR is the registered trademark of HL7 and is used with the permission of HL7. Use of the FHIR trademark does not constitute endorsement of this product by HL7] resources.

FHIR stands for *Fast Healthcare Interoperability Resources* and  is a standard describing data formats and elements (known as "resources") as well as an application programming interface (API) for exchanging electronic health records. The standard was created by the Health Level Seven International (HL7) health-care standards organization. For more information on the FHIR standard, visit https://www.hl7.org/fhir/.

While FHIR is a very useful standard to describe and exchange medical data in an interoperable way, it is not at all useful for statistical analyses of data. This is due to the fact that FHIR data is stored in many nested and interlinked resources instead of matrix-like structures.

Thus, to be able to do statistical analyses a tool is needed that allows converting these nested resources into data frames. This process of tabulating FHIR resources is not trivial, as the unpredictable degree of nesting and connectedness of the resources makes generic solutions to this problem not feasible.

We therefore implemented a package that makes it possible to download FHIR resources from a server into R and to tabulate these resources into (multiple) data frames.

The package is still under development. The CRAN version of the package contains all functions that are already stable, for more recent (but potentially unstable) developments, the development version of the package can be downloaded from GitHub using `devtools::install_github("POLAR-fhiR/fhircrackr")`.

This vignette is an introduction on the basic functionalities of the `fhircrackr` and should give you a broad overview over what the package can do. For more detailed instructions on each subtopic please have a look the other vignettes. This introduction covers the following topics:

- Prerequisites

- Downloading resources from a FHIR server

- Flattening resources

- Multiple entries

- Saving and loading downloaded bundles

- Creating resources


## Prerequisites
The complexity of the problem requires a couple of prerequisites both regarding your knowledge and access to data. We will shortly list the  preconditions for using the `fhircrackr` package here:  

1. First of all, you need the base URL of the FHIR server you want to access. If you don't have your own FHIR server, you can use one of the available public servers, such as `https://hapi.fhir.org/baseR4` or `http://fhir.hl7.de:8080/baseDstu3`. The base URL of a FHIR server is often referred to as [base].

2. To download resources from the server, you should be familiar with [FHIR search requests](https://www.hl7.org/fhir/search.html). FHIR search allows you to download sets of resources that match very specific requirements. The `fhircrackr` package offers some help building FHIR search requests, for this please see the vignette on downloading FHIR resources.

3. In the first step, `fhircrackr` downloads the resources in xml format into R. To specify which elements from the FHIR resources you want in your data frame, you should have at least some familiarity with XPath expressions. A good tutorial on XPath expressions can be found [here](https://www.w3schools.com/xml/xpath_intro.asp).

In the following we'll go through a typical workflow with `fhircrackr` step by step. The first and foremost step is of course, to install and load the package:

```{r, eval=F}
install.packages("fhircrackr")
library(fhircrackr)
```

```{r, include=F}
library(fhircrackr)
```

## Downloading resources
To download resources from a FHIR server, you need to send a FHIR search request using `fhir_search()`.
This introduction will not go into the details of building a valid FHIR search request. For that, please see the vignette on downloading FHIR resources. Here we will use a simple example of downloading all Patient resources from a public HAPI server:

```{r, eval=F}
patient_bundles <- fhir_search(request="http://fhir.hl7.de:8080/baseDstu3/Patient",
							   max_bundles=2, verbose = 0)

```


```{r, include=F}
patient_bundles <- fhir_unserialize(fhircrackr::patient_bundles)
``` 

The minimum information `fhir_search()` requires is a string containing the full FHIR search request in the argument `request` which you can either provide explicitly or by a call to `fhir_build_url()` before.
In general, a FHIR search request returns a *bundle* of the resources you requested. If there are a lot of resources matching your request, the search result isn't returned in one big bundle but distributed over several of them. If the argument `max_bundles` is set to its default `Inf`, `fhir_search()` will return all available bundles, meaning all resources matching your request. If you set it to `2` as in the example above, the download will stop after the first two bundles. Note that in this case, the result *may not contain all* the resources from the server matching your request.

If you want to connect to a FHIR server that uses basic authentication, you can supply the arguments `username` and `password`.

As you can see in the next block of code, `fhir_search()` returns a list of xml objects where each list element represents one bundle of resources, so a list of two xml objects in our case:

```{r}
length(patient_bundles)
str(patient_bundles[[1]])
```

If for some reason you cannot connect to a FHIR server at the moment but want to explore the following functions anyway, the package provides two example lists of bundles containing Patient and MedicationStatement resources. See `?patient_bundles` and `?medication_bundles` for how to use them.


## Flattening resources
Now we know that inside these xml objects there is the patient data somewhere. To get it out, we will use `fhir_crack()`. The most important argument `fhir_crack()` takes is `bundles`, the list of bundles that is returned by `fhir_search()`. The second important argument is `design`, an object that tells the function which data to extract from the bundle. `fhir_crack()` returns a list of data.frames (the default) or a list of data.tables (if argument `data.tables=TRUE`).

The proper format of a `design` in `fhir_crack()` is described in detail in the vignette on flattening resources. Please refer to this document for more information, as we will just use one example of a design here.

In general, `design` is a named list containing one element per data frame that will be created. We call these elements *data.frame descriptions*. The names of the data.frame descriptions in `design` are also going to be the names of the resulting data frames. It usually makes sense to create one data frame per type of resource. Because we have just downloaded resources of the type Patient, the `design` here would be a list of length 1, containing just one data.frame description. A full data.frame description is a list with the elements *resource*, *cols* and *style* and can look as follows:

```{r}
#define design
design <- list(

	Patients = list(
		
		resource = "//Patient",
		
		cols = list(
			PID           = "id",
			use_name      = "name/use",
			given_name    = "name/given",
			family_name   = "name/family",
			gender        = "gender",
			birthday      = "birthDate"
		),
		
		style = list(
			sep = "|",
			brackets = c("[","]"),
			rm_empty_cols = FALSE
		)
	)
)
```

All three elements of `style` can also be controlled directly by the `fhir_crack()` arguments `sep`, `brackets` and `remove_empty_columns`. If the function arguments are `NULL` (their default), the values provided in `style` are used, if they are not NULL, they will overwrite any values in `style`. If both the function arguments and the `style` component of the data.frame description are NULL, default values(`sep=" "`, `brackets = NULL`, `rm_empty_cols=TRUE`) will be assumed. 

After it is defined, the design can be used in `fhir_crack()` like this:

```{r}
#flatten resources
list_of_tables <- fhir_crack(bundles = patient_bundles, design = design, verbose = 0)

#have look at the results
head(list_of_tables$Patients)
```